---
title: "LDA_IGA-NTX"
author: "Sven Morlock"
date: "28/03/2022"
header-includes:
  - \usepackage{xspace}
  - \usepackage{tikz}
  - \usetikzlibrary{shapes.geometric,arrows}
  
output: pdf_document
---
```{r setup, include = FALSE}
#knitr::opts_chunck$set(fig.align = "left")
```

# § 1

Load required packages:

```{r, message = FALSE, echo = FALSE}
library(ggplot2)
library(patchwork)
library(kableExtra)
library(survival)
library(survminer)
library(data.table)
library(readxl)
library(forcats)
library(lubridate)
library(dplyr)
```

```{r}
#if (Sys.info()["sysname"] != "Darwin"){
#  default_theme <- theme_minimal() +
#  theme(text = element_text(family = "Decima WE", size = 15)) +
#  theme(panel.grid.major = element_line(color = "grey", size = 0.3)) +
#  theme(axis.line = element_line(color = "black", size = 0.4))
#
#} else {
#  default_theme <- theme_minimal()
#}
default_theme <- theme_minimal()
two_scale_fill <- scale_fill_manual(values=c("#69b3a2", "#404080"))
```

Read data using read_data.R scrip:

```{r,, message = FALSE}
source("read_data.R")
```
\newpage
---
# § 2 EDA

```{r}
follow_up <- years(10)
data_iga$follow_up_truncated <-pmin(data_iga$`T-dls`, data_iga$`T-date` + follow_up)

data_iga$mismatch_sum <- as.numeric(as.character(data_iga$`mm-A`)) + as.numeric(as.character(data_iga$`mm-B`)) + as.numeric(as.character(data_iga$`mm-DR`))

data_iga_pos <- data_iga[`biopsy proven recurrence (0=no, 1=yes)` == 1]
data_iga_neg <- data_iga[`biopsy proven recurrence (0=no, 1=yes)` == 0]
```

## § 2.1 IGA
<!-- ################################## FOLLOW UP #####################  -->
```{r}
# follow-up mean / median 
tbl_iga_follow_up_mean <- data.frame(
  iga_all = mean((interval(data_iga$`T-date`, data_iga$`T-dls`) / years(1))),
  iga_pos = mean((interval(data_iga_pos$`T-date`, data_iga_pos$`T-dls`) / years(1))),
  iga_neg = mean((interval(data_iga_neg$`T-date`, data_iga_neg$`T-dls`) / years(1)))
)
tbl_iga_follow_up_median <- data.frame(
  iga_all = median((interval(data_iga$`T-date`, data_iga$`T-dls`) / years(1))),
  iga_pos = median((interval(data_iga_pos$`T-date`, data_iga_pos$`T-dls`) / years(1))),
  iga_neg = median((interval(data_iga_neg$`T-date`, data_iga_neg$`T-dls`) / years(1)))
)
```

```{r}
p1 <- ggplot() +
  geom_boxplot(aes(x = "", y = interval(data_iga$`T-date`, data_iga$`T-dls`) / years(1)), data = data_iga) +
                 default_theme +
  ylab("Jahre") +
  xlab("")

p2 <- ggplot() +
  geom_boxplot(aes(x = `biopsy proven recurrence (0=no, 1=yes)`, y = interval(data_iga$`T-date`, data_iga$`T-dls`) / years(1)), data = data_iga) +
  default_theme +
  ylab("Jahre") +
  xlab("") +
  scale_x_discrete(labels = c("IgA-", "IgA+"))
  
patch <- p1 | p2

patch
```


```{r}
# follow_up mean/median (truncated)
tbl_iga_follow_up_mean_truncated <- data.frame(
  iga_all = mean((interval(data_iga$`T-date`, data_iga$follow_up_truncated) / years(1))),
  iga_pos = mean((interval(data_iga_pos$`T-date`, data_iga_pos$follow_up_truncated) / years(1))),
  iga_neg = mean((interval(data_iga_neg$`T-date`, data_iga_neg$follow_up_truncated) / years(1)))
)
tbl_iga_follow_up_median_truncated <- data.frame(
  iga_all = median((interval(data_iga$`T-date`, data_iga$follow_up_truncated) / years(1))),
  iga_pos = median((interval(data_iga_pos$`T-date`, data_iga_pos$follow_up_truncated) / years(1))),
  iga_neg = median((interval(data_iga_neg$`T-date`, data_iga_neg$follow_up_truncated) / years(1)))
)
```

<!-- ################################## DEATH WITHIN FOLLOW UP #####################  -->
```{r}
# patient death within follow up
tbl_iga_pat_death <- data.frame(
  iga_all = nrow(data_iga[(`T-dls` <= (`T-date` + follow_up)) & `Pat death (0=alive, 1= dead)` == 1]),
  iga_pos = nrow(data_iga_pos[(`T-dls` <= (`T-date` + follow_up)) & `Pat death (0=alive, 1= dead)` == 1]),
  iga_neg = nrow(data_iga_neg[(`T-dls` <= (`T-date` + follow_up)) & `Pat death (0=alive, 1= dead)` == 1])
)
```
<!-- ################################## DROP OUT WITHIN FOLLOW UP #####################  -->
```{r}
# patient drop out within follow up
tbl_iga_pat_drop <- data.frame(
  iga_all = nrow(data_iga[(`T-dls` <= (`T-date` + follow_up)) & `Pat death (0=alive, 1= dead)` == 0]),
  iga_pos = nrow(data_iga_pos[(`T-dls` <= (`T-date` + follow_up)) & `Pat death (0=alive, 1= dead)` == 0]),
  iga_neg = nrow(data_iga_neg[(`T-dls` <= (`T-date` + follow_up)) & `Pat death (0=alive, 1= dead)` == 0])
)
```

<!-- ################################## GRAFT LOSS           #####################  -->
```{r}
# patients with graft loss 
tbl_iga_graft_loss <- data.frame(
  iga_all = nrow(data_iga[`graft loss (0=functial, 1=loss)` == 1]),
  iga_pos = nrow(data_iga_pos[`graft loss (0=functial, 1=loss)` == 1]),
  iga_neg = nrow(data_iga_neg[`graft loss (0=functial, 1=loss)` == 1])
)
```

```{r}
# patients with graft loss within follow up period
tbl_iga_graft_loss_follow_up <- data.frame(
  iga_all = nrow(data_iga[`graft loss date` < `T-date` + follow_up]),
  iga_pos = nrow(data_iga_pos[`graft loss date` < `T-date` + follow_up]),
  iga_neg = nrow(data_iga_neg[`graft loss date` < `T-date` + follow_up])
)
```
<!-- ############################################ AGE #################################  -->
```{r}
# ToDo: median IQR
# age patients (yrs.)
ggplot(data = data_iga) +
  geom_histogram(mapping = aes(x = interval(`Date of birth`,`T-date`) / years(1),
                               fill = `R-sex`),
                 color="#e9ecef",
                 alpha=0.6,
                 bins = 40,
                 position = "dodge") +
  two_scale_fill +
  ylab("Anzahl") +
  xlab("Alter in Jahren") +
  ggtitle("Histogram Alter in Jahren am Transplantationsdatum",
          subtitle =  "IGA all") +
  labs(fill = "Geschlecht") +
  default_theme
```
```{r}
ggplot(data = data_iga_pos) +
 geom_histogram(mapping = aes(x = interval(`Date of birth`,`T-date`) / years(1),
                              fill = `R-sex`),
                color="#e9ecef",
                alpha=0.6,
                bins = 40,
                position = "dodge") +
 two_scale_fill +
 ylab("Anzahl") +
 xlab("Alter in Jahren") +
 ggtitle("Histogram Alter in Jahren am Transplantationsdatum",
         subtitle =  "IGA+") +
 labs(fill = "Geschlecht") +
 default_theme
```
```{r}
ggplot(data = data_iga_neg) +
 geom_histogram(mapping = aes(x = interval(`Date of birth`,`T-date`) / years(1),
                              fill = `R-sex`),
                color="#e9ecef",
                alpha=0.6,
                bins = 40,
                position = "dodge") +
 two_scale_fill +
 ylab("Anzahl") +
 xlab("Alter in Jahren") +
 ggtitle("Histogram Alter in Jahren am Transplantationsdatum",
         subtitle =  "IGA-") +
 labs(fill = "Geschlecht") +
 default_theme
```

```{r}
## IQR
tbl_iga_iqr_age <- data.frame(
  iga_all = IQR(data_iga$R_age_Tdate),
  iga_po = IQR(data_iga_pos$R_age_Tdate),
  iga_neg = IQR(data_iga_neg$R_age_Tdate)
)
```

```{r}
## boxpot

p1 <- ggplot(data = data_iga) +
  geom_boxplot(aes(x = R_age_Tdate)) +
  coord_flip() +
  ggtitle("IGA (all)") +
  default_theme 

p2 <- ggplot(data = data_iga_pos) +
  geom_boxplot(aes(x = R_age_Tdate)) +
  coord_flip() +
  ggtitle("IGA r+") +
  default_theme 

p3 <- ggplot(data = data_iga_neg) +
  geom_boxplot(aes(x = R_age_Tdate)) +
  coord_flip()+
  ggtitle("IGA r-") +
  default_theme

p1$labels$x <- p2$labels$x <- p3$labels$x <- ""
p1$labels$y<- p2$labels$y <- p3$labels$y <- ""

patch <- p1 | p2 | p3
patch <- patch + plot_annotation(
  title = "Boxplot Altersverteilung"
)
patch & xlim(0, 80) & theme(axis.ticks.x = element_blank(),
                            axis.text.x = element_blank()) & xlab("Alter")
```

```{r}
# sex
tbl_iga_sex_abs <- data.frame(
  iga_all = summary(data_iga$`R-sex`),
  iga_pos = summary(data_iga_pos$`R-sex`),
  iga_neg = summary(data_iga_neg$`R-sex`)
)
```

```{r}
p1 <- data.frame(group = rownames(tbl_iga_sex_abs),
           value = tbl_iga_sex_abs$iga_all) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Geschlecht")) +
  theme_void() +
  two_scale_fill +
  ggtitle("iga all")
  
p2 <- data.frame(group = rownames(tbl_iga_sex_abs),
           value = tbl_iga_sex_abs$iga_pos) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Geschlecht")) +
  theme_void() +
  two_scale_fill +
  ggtitle("iga +")

p3 <- data.frame(group = rownames(tbl_iga_sex_abs),
           value = tbl_iga_sex_abs$iga_neg) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Geschlecht")) +
  theme_void() +
  two_scale_fill +
  ggtitle("iga -")

patch <- p1 | p2 | p3
patch + plot_annotation(title = "Kreisdiagramme Geschlecht (absolut)")
```

```{r}
# sex percent
tbl_iga_sex_percent <- data.frame(
  iga_all = round(summary(data_iga$`R-sex`) / nrow(data_iga), 2),
  iga_pos = round(summary(data_iga_pos$`R-sex`) / nrow(data_iga_pos), 2),
  iga_neg = round(summary(data_iga_neg$`R-sex`) / nrow(data_iga_neg), 2)
)
```


```{r}
# death donator
tbl_death_d <- cbind(c(nrow(data_iga[`Pat death (0=alive, 1= dead)` == 0]),
                       nrow(data_iga[`Pat death (0=alive, 1= dead)` == 1])),
                     c(nrow(data_iga_pos[`Pat death (0=alive, 1= dead)` == 0]),
                       nrow(data_iga_pos[`Pat death (0=alive, 1= dead)` == 1])),
                     c(nrow(data_iga_neg[`Pat death (0=alive, 1= dead)` == 0]),
                       nrow(data_iga_neg[`Pat death (0=alive, 1= dead)` == 1])))
colnames(tbl_death_d) <- c("iga_all", "iga_pos", "iga_neg")
rownames(tbl_death_d) <- c("Lebendspende", "Todspende")

p1 <- data.frame(group = c("Lebendspende", "Todspende"),
           value = c(nrow(data_iga[`Pat death (0=alive, 1= dead)` == 0]),
                      nrow(data_iga[`Pat death (0=alive, 1= dead)` == 1]))) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Spender")) +
  theme_void() +
  two_scale_fill +
  ggtitle("iga all")

p2 <- data.frame(group = c("Lebendspende", "Todspende"),
           value = c(nrow(data_iga_pos[`Pat death (0=alive, 1= dead)` == 0]),
                      nrow(data_iga_pos[`Pat death (0=alive, 1= dead)` == 1]))) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Spender")) +
  theme_void() +
  two_scale_fill +
  ggtitle("iga +")

p3 <- data.frame(group = c("Lebendspende", "Todspende"),
           value = c(nrow(data_iga_neg[`Pat death (0=alive, 1= dead)` == 0]),
                      nrow(data_iga_neg[`Pat death (0=alive, 1= dead)` == 1]))) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Spender")) +
  theme_void() +
  two_scale_fill +
  ggtitle("iga -")


patch <- p1 | p2 | p3
patch + plot_annotation(title = "Kreisdiagramme Spender")
  
```

```{r}
# ToDo: iga all + -
# deceased D.
# living D.
tbl_iga_1 <- summary(data_iga[`graft loss (0=functial, 1=loss)` == 1]$`D-type`)
tbl_iga_1 <- round(tbl_iga_1 / sum(tbl_iga_1), 3)
tbl_iga_2 <- summary(data_iga[`graft loss (0=functial, 1=loss)` == 0]$`D-type`)
tbl_iga_2 <- round(tbl_iga_2 / sum(tbl_iga_2),3)
rbind("loss" = tbl_iga_1, "functional" = tbl_iga_2) %>% 
  as.data.frame()
```

```{r}
# dead/alive
tbl_iga_1 <- summary(data_iga[`Pat death (0=alive, 1= dead)` == 1]$`D-type`)
tbl_iga_1 <- round(tbl_iga_1 / sum(tbl_iga_1), 3)
tbl_iga_2<- summary(data_iga[`Pat death (0=alive, 1= dead)` == 0]$`D-type`)
tbl_iga_2 <-  round(tbl_iga_2 / sum(tbl_iga_2),3)  # in %
rbind("dead" = tbl_iga_1, "alive" = tbl_iga_2) %>% 
  as.data.frame()
```


```{r}
# BMI (mean.)
tbl_iga_bmi <- data.table(
  iga_all = mean(data_iga$`D-weight` * (data_iga$`D-height`)^2),
  iga_pos = mean(data_iga_pos$`D-weight` * (data_iga_pos$`D-height`)^2),
  iga_neg = mean(data_iga_neg$`D-weight` * (data_iga_neg$`D-height`)^2)
)
```


```{r}
p1 <- data_iga %>% 
  ggplot() +
  geom_boxplot(aes(x = "", y = mismatch_sum )) +
  default_theme +
  ggtitle("IGA all") +
  ylab("") +
  xlab("")

p2 <- data_iga_neg %>% 
  ggplot() +
  geom_boxplot(aes(x = "", y = mismatch_sum)) +
  default_theme +
  ggtitle("IGA-") +
  ylab("") +
  xlab("")

p3 <- data_iga_pos %>% 
  ggplot() +
  geom_boxplot(aes(x = "", y = mismatch_sum)) +
  default_theme +
  ggtitle("IGA+") + 
  ylab("") +
  xlab("")

patch <- p1 | p2 | p3
patch + plot_annotation(title = "HLA mismatch summe")

```


```{r}
# HLA-mm (0-6)
## data.table of mm-A, mm-B and mm-DR
tbl_iga_1 <- as.data.table(lapply(data_iga[, c("mm-A", "mm-B", "mm-DR")],
                              as.numeric))
## calculate mean for each column
tbl_iga_2 <- apply(X = tbl_iga_1,
               MARGIN = 2,
               FUN = mean, na.rm = TRUE)
round(tbl_iga_2, 3)
# mean of means
round(mean(tbl_iga_2),3)

## sum of mm-A, mm-B and mm-DR
tbl_iga_3 <- apply(X = tbl_iga_1,
               MARGIN = 1,  
               FUN = sum, na.rm = TRUE)
# mean value of col sum
mean(tbl_iga_3)
# median value of col sum
sd(tbl_iga_3)
```

```{r}
# PRA current (mean)
tbl_iga_pra_curr_mean <- data.frame(
  iga_all = mean(data_iga$`Current PRA%`, na.rm = TRUE),
  iga_pos = mean(data_iga_pos$`Current PRA%`, na.rm = TRUE),
  iga_neg = mean(data_iga_neg$`Current PRA%`, na.rm = TRUE)
)
# PRA highest (mean)
tbl_iga_pra_high <- data.frame(
  iga_all = mean(data_iga$`Highest PRA%`, na.rm = TRUE),
  iga_pos = mean(data_iga_pos$`Highest PRA%`, na.rm = TRUE),
  iga_neg = mean(data_iga_neg$`Highest PRA%`, na.rm = TRUE)
)

```

```{r}
# age donor (mean.)
tbl_iga_age_donor <- data.frame(
  iga_all = mean(data_iga$`D-age`),
  iga_pos =  mean(data_iga_pos$`D-age`),
  iga_neg =  mean(data_iga_neg$`D-age`)
)
```

```{r}
# cold-ischemia time (hours)
# mean
tbl_iga_cis_mean <- data.frame(
  iga_all = mean(data_iga$`Cold ischaemic period hours`, na.rm = TRUE),
  iga_pos = mean(data_iga_pos$`Cold ischaemic period hours`, na.rm = TRUE),
  iga_neg = mean(data_iga_neg$`Cold ischaemic period hours`, na.rm = TRUE)
)
# median
tbl_iga_cis_median <- data.frame(
  iga_all = median(data_iga$`Cold ischaemic period hours`, na.rm = TRUE),
  iga_pos = median(data_iga_pos$`Cold ischaemic period hours`, na.rm = TRUE),
  iga_neg = median(data_iga_neg$`Cold ischaemic period hours`, na.rm = TRUE)
)
# standard error
tbl_iga_cis_sd <- data.frame(
  iga_all = sd(data_iga$`Cold ischaemic period hours`, na.rm = TRUE),
  iga_pos = sd(data_iga_pos$`Cold ischaemic period hours`, na.rm = TRUE),
  iga_neg = sd(data_iga_neg$`Cold ischaemic period hours`, na.rm = TRUE)
)
```

```{r}
p1 <- data_iga %>% 
ggplot() +
geom_boxplot(aes(x = `Cold ischaemic period hours`)) +
  coord_flip() +
  ggtitle("IGA (all)") +
  default_theme 

p2 <- data_iga_pos %>% 
  ggplot() +
  geom_boxplot(aes(x = `Cold ischaemic period hours`)) +
  coord_flip() +
  ggtitle("IGA +") +
  default_theme 

p3 <- data_iga_neg %>% 
  ggplot() +
  geom_boxplot(aes(x = `Cold ischaemic period hours`)) +
  coord_flip() +
  ggtitle("IGA -") +
  default_theme 

p1$labels$x <- p2$labels$x <- p3$labels$x <- ""
p1$labels$y<- p2$labels$y <- p3$labels$y <- ""

patch <- p1 | p2 | p3
patch <- patch + plot_annotation(
  title = "Boxplot Cold Isch. Time h"
)
patch & xlim(0, 40) & theme(axis.ticks.x = element_blank(),
                           axis.text.x = element_blank()) & xlab("Cold Isch Time")
```



```{r}
# living vs dead donator
tbl_iga_don_living_abs <- data.frame(
  iga_all = nrow(data_iga[`D-type` == "Living"]),
  iga_pos = nrow(data_iga_pos[`D-type` == "Living"]),
  iga_neg = nrow(data_iga_neg[`D-type` == "Living"])
)

tbl_iga_don_dead_abs <- data.frame(
  iga_all = nrow(data_iga[`D-type` == "Cadaver"]),
  iga_pos = nrow(data_iga_pos[`D-type` == "Cadaver"]),
  iga_neg = nrow(data_iga_neg[`D-type` == "Cadaver"])
)


tbl_iga_don_living_rel <- data.frame(
  iga_all = nrow(data_iga[`D-type` == "Living"]) / nrow(data_iga),
  iga_pos = nrow(data_iga_pos[`D-type` == "Living"]) / nrow(data_iga_pos),
  iga_neg = nrow(data_iga_neg[`D-type` == "Living"]) / nrow(data_iga_neg)
)

tbl_iga_don_dead_rel <- data.frame(
  iga_all = nrow(data_iga[`D-type` == "Cadaver"]) / nrow(data_iga),
  iga_pos = nrow(data_iga_pos[`D-type` == "Cadaver"]) / nrow(data_iga_pos),
  iga_neg = nrow(data_iga_neg[`D-type` == "Cadaver"]) / nrow(data_iga_neg)
)
```


```{r}
# Überleben MEdian IQR
tbl_iga_surv_median <- data.frame(
  iga_all = median(data_iga[`Pat death (0=alive, 1= dead)` == 1, R_age_Tdls], na.rm = TRUE),
  iga_pos = median(data_iga_pos[`Pat death (0=alive, 1= dead)` == 1, R_age_Tdls], na.rm = TRUE),
  iga_neg = median(data_iga_neg[`Pat death (0=alive, 1= dead)` == 1, R_age_Tdls], na.rm = TRUE))

tbl_iga_surv_iqr <- data.frame(
  iga_all = IQR(data_iga[`Pat death (0=alive, 1= dead)` == 1, R_age_Tdls], na.rm = TRUE),
  iga_pos = IQR(data_iga_pos[`Pat death (0=alive, 1= dead)` == 1, R_age_Tdls], na.rm = TRUE),
  iga_neg = IQR(data_iga_neg[`Pat death (0=alive, 1= dead)` == 1, R_age_Tdls], na.rm = TRUE)
)

```

```{r}
iga_table <- rbindlist(
  list(
    iga_dropout = data.frame(
      cbind(name = "dropout",tbl_iga_pat_drop)),
    iga_graftloss = data.frame(
      cbind(name = "graftloss", tbl_iga_graft_loss)),
    iga_graftloss_follow_up = data.frame(
      cbind(name = "graftloss_followup", tbl_iga_graft_loss_follow_up)),
    iga_pat_death = data.frame(
      cbind(name = "pat_death", tbl_iga_pat_death)),
    iga_don_dead_abs = data.frame(
      cbind(name = "don_dead_abs", tbl_iga_don_dead_abs)),
    iga_don_dead_rel = data.frame(
      cbind(name = "don_dead_rel", tbl_iga_don_dead_rel)),
    iga_don_living_abs = data.frame(
      cbind(name = "don_living_abs", tbl_iga_don_living_abs)),
    iga_don_living_rel = data.frame(
      cbind(name = "don_living_rel", tbl_iga_don_living_rel)),
    iga_cis_h_mean = data.frame(
      cbind(name = "cold_h_mean", tbl_iga_cis_mean)),
    iga_cis_h_sd = data.frame(
      cbind(name = "cold_h_sd", tbl_iga_cis_sd)),
    iga_median_survtime = data.frame(
      cbind(name = "median survival time", tbl_iga_surv_median)),
    iga_iqr_survtime = data.frame(
      cbind(name = "IQR survival time", tbl_iga_surv_iqr))
    ))

kable(iga_table, col.names = c("Merkmal", "IgA (all)", "IgA r+", "IgA r-")) %>% 
  kable_styling(latex_options = "hold_position")

```


## § 2.2 NTX

```{r}
# follow_up mean
pmin(
  # follow up
  interval(data_ntx$Datum, (data_ntx$Datum + follow_up)) / years(1),
  # last seen
  interval(data_ntx$Datum, data_ntx$tdls) / years(1)
) %>%
  mean()
```

```{r}
ggplot() +
  geom_boxplot(aes(x = "",
                   y = interval(data_ntx$Datum, data_ntx$tdls) / years(1)), data = data_ntx) +
  default_theme +
  ggtitle("Boxplot: Follow Up Period") +
  ylab("Jahre") +
  xlab("") 
```

```{r}
# patient death within follow up
nrow(data_ntx[`Todesdatum[NTX PatientenInformation]` < (Datum + follow_up)])
```

```{r}
# patient drop out
nrow(data_ntx[`Date last seen[NTX PatientenInformation]` <= (Datum + follow_up) & `Patienten Status[NTX PatientenInformation]` == "1 - lebt"])
```

```{r}
# patients with graft loss
nrow(data_ntx[data_ntx$`TX Status[NTX PatientenInformation]` == "2 - ohne Transplantatfunktion"])
```

```{r}
# age patients (yrs.)
ggplot(data = data_ntx) +
  geom_histogram(mapping = aes(x = R_age_Datum, fill = Geschlecht),
                 color = "#e9ecef", alpha = 0.6, position = "dodge") +
  two_scale_fill +
  ylab("Anzahl") +
  xlab("Alter in Jahren") +
  ggtitle("Histogram Alter in Jahren am Transplantationsdatum") +
  labs(fill = "Geschlecht") +
  default_theme
```
```{r}
ggplot(data = data_ntx) +
  geom_boxplot(aes(x = data_ntx$Geschlecht, y = data_ntx$R_age_Datum))  +
  default_theme +
  ggtitle("Boxplot Alter NTX") +
  xlab("Geschlecht") + 
  ylab("Alter")

```


```{r}
# male sex
summary(data_ntx$Geschlecht)
```

```{r}
data.frame(group = names(summary(data_ntx$Geschlecht)),
              value = c(summary(data_ntx$Geschlecht))) %>% 
  ggplot(aes(x = "", y = value, fill = group)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  guides(fill=guide_legend(title = "Geschlecht")) +
  theme_void() +
  two_scale_fill +
   ggtitle("NTX")
```

```{r}
median(interval(data_ntx$Geburtsdatum, data_ntx$`Todesdatum[NTX PatientenInformation]`) / years(1), na.rm = TRUE)
```

```{r}
IQR(interval(data_ntx$Geburtsdatum, data_ntx$`Todesdatum[NTX PatientenInformation]`) / years(1), na.rm = TRUE)
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```


\newpage
---
# § 3 Kaplan-Meier


## § 3.1 IGA

```{r}
#  functions in survival and survminer package need numeric-type input
data_iga[, time_date_biopsy := interval(`T-date`, `date of biopsy`) / years(1)]
data_iga[, time_t_dls := interval(`T-date`, `T-dls`) / years(1)]
data_iga[, time_date_birth := interval(`T-date`, `Date of birth`) / years(1)]
data_iga[, time_graft_loss := interval(`T-date`, `graft loss date`) / years(1)]
data_iga[, time_date_follow_up := interval(`T-date`,`T-date` + follow_up) / years(1)]
```
### § 3.1.1 

- Event:
  - graft-loss within the follow up period.
  
- Censoring scheme:
  - if `graft loss date` after follow up period, censored by end of follow up.
  - if `T-dls` (date last seen) within follow up period, censored by `T-dls`.
  
- Time period:
  -  10 years after `T-date` (kidney transplantaion).
```{r}
data_iga <- data_iga %>% 
  mutate(status_date = case_when(
    ## graft-loss within follow up period
    !is.na(`graft loss date`) & `graft loss date` <  `T-date` + follow_up ~ time_graft_loss,
    ## graft-loss after follow up period
    !is.na(`graft loss date`) & `graft loss date` >  `T-date` + follow_up ~ time_date_follow_up,
    ## no graft-loss and last seen within follow up
    is.na(`graft loss date`) & !is.na(`T-dls`) & `T-dls` < `T-date` + follow_up ~ time_t_dls,
    ## no graft-loss and last seen after follow up
    is.na(`graft loss date`) & !is.na(`T-dls`) & `T-dls` > `T-date` + follow_up ~ time_date_follow_up,
    ## no graft loss and no last seen 
    is.na(`graft loss date`) & is.na(`T-dls`) ~ time_date_follow_up
                                )
  )
data_iga <- data_iga %>% 
  mutate(status = case_when(
    ## graft-loss within follow up period
    !is.na(`graft loss date`) & `graft loss date` <  `T-date` + follow_up ~ 1,
    ## else censored
    TRUE ~ 0,
                                )
  )
data_iga_pos <- data_iga[`biopsy proven recurrence (0=no, 1=yes)` == 1]
data_iga_neg <- data_iga[`biopsy proven recurrence (0=no, 1=yes)` == 0]
```

##### Overall kaplan-Meier curve


```{r}
model_iga_1 <- survfit(formula = Surv(time = status_date, event = status, type = "right") ~ data_iga$`biopsy proven recurrence (0=no, 1=yes)`, data_iga)

ggsurvplot(model_iga_1)
```

```{r}
model_iga_1.1 <- survfit(formula = Surv(time = status_date,
                                   event = status, type = "right")~ 1,
                    data = data_iga)  
ggsurvplot(model_iga_1.1,
           conf.int = FALSE,
           cumevents = TRUE)
```


```{r}
## iga+
model_iga_1.1 <- survfit(formula = Surv(time = status_date,
                                   event = status, type = "right")~ 1,
                    data = data_iga_pos)  
ggsurvplot(model_iga_1.1,
           conf.int = FALSE,
           cumevents = TRUE)

```

```{r}
## iga-
model_iga_1.2 <- survfit(formula = Surv(time = status_date,
                                   event = status, type = "right")~ 1,
                    data = data_iga_neg)  
ggsurvplot(model_iga_1.2,
           conf.int = FALSE,
           cumevents = TRUE)

```

### § 3.1.2
- Event: 
  - `T-dls` & ``Pat death (0=alive, 1= dead)` (patient death).
- Censoring scheme:
  - `T-dls` (date last seen) within follow up period.
- Time period:
  - 10 years after `T-date` (kidney transplantation).
```{r}
data_iga <- data_iga %>% 
  mutate(status_date = case_when(
    ## patient death and death date within follow up
    (`Pat death (0=alive, 1= dead)` == 1) & `T-dls` < `T-date` + follow_up ~ time_t_dls,
    ## patient dead but after follow up
    (`Pat death (0=alive, 1= dead)` == 1) & `T-dls` > `T-date` + follow_up ~ time_date_follow_up,
    ## patient not death but dropped within follow up
    (`Pat death (0=alive, 1= dead)` == 0) & `T-dls` < `T-date` + follow_up ~ time_t_dls,
    ## patient not death but dropped after follow up
    (`Pat death (0=alive, 1= dead)` == 0) & `T-dls` > `T-date` + follow_up ~ time_date_follow_up,
    ## NOTE: T-dls never NA
  )
)
```

```{r}
model_iga_2 <- survfit(formula = Surv(time = status_date,
                                   event = status, type = "right") ~ data_iga$`biopsy proven recurrence (0=no, 1=yes)`,
                    data = data_iga)  
ggsurvplot(model_iga_2,
           conf.int = FALSE)
```

ASDFGHJKLKJHGFDSASDFGHJKLKJHGFDS

#### § 3.1.3
- Event: graft loss and death
- Censoring scheme:
- Time period:

\newpage
## § 3.2 IGA

### § 3.2.1
- Event: `TX Status[NTX PatientenInformation]`
- Censoring scheme: 
  -if `TX Status[NTX PatientenInformation]` within follow up period then then event
  - if died within follow up  and before no graft loss then censored
  - if last seen within follow up and before no graft loss then censored
  - if graft loss after follow up then censored
- Time period: 10 years after `Datum`

```{r}
data_ntx <- data_ntx %>%
  mutate(status_date = case_when(
    ## patient experienced graft loss
    !is.na(Transplantatfunktionsende) & Transplantatfunktionsende <= (Datum + follow_up) ~ interval(Datum, Transplantatfunktionsende) / years(1),
    ## patient died within follow up
    `Todesdatum[NTX PatientenInformation]` < (Datum + follow_up) ~ interval(Datum, `Todesdatum[NTX PatientenInformation]`) / years(1),
    ## patiend last seen within follow up
    `Date last seen[NTX PatientenInformation]` < (Datum + follow_up) ~ interval(Datum, `Date last seen[NTX PatientenInformation]`) / years(1),
    ## else follow up
    TRUE ~ interval(Datum, (Datum + follow_up)) / years(1)
    )
  )

data_ntx <- data_ntx %>%
  mutate(status = case_when(
    ## patient experienced graft loss
    !is.na(Transplantatfunktionsende) & Transplantatfunktionsende <= (Datum + follow_up) ~ 1,
    ## patient died within follow up 
    `Todesdatum[NTX PatientenInformation]` < (Datum + follow_up) ~ 0,
    ## patiend last seen within follow up
    `Date last seen[NTX PatientenInformation]` < (Datum + follow_up) ~ 0,
    ## else follow up
    TRUE ~ 0
    )
  )
```

```{r}
model_ntx_1 <- survfit(formula = Surv(time = status_date,
                                      event = status, type = "right") ~ 1,
                       data = data_ntx)

ggsurvplot(model_ntx_1,
           conf.int = FALSE,
           cumevents = TRUE)
```

### § 3.2.2

- Event: patient died within follow up period
- Censoring scheme:
  - patient died within follow up then event
  - patient dropped from study within  follow up then censored

```{r}
data_ntx <- data_ntx %>%
  mutate(status_date = case_when(
    ## patient died within follow up
    `Todesdatum[NTX PatientenInformation]` <= (Datum + follow_up) ~ interval(Datum, `Todesdatum[NTX PatientenInformation]`) / years(1),
    ## patient died after follow up
    `Todesdatum[NTX PatientenInformation]` > (Datum + follow_up) ~ interval(Datum, (Datum + follow_up)) / years(1),
    ## patient dropped within follow up
    `Date last seen[NTX PatientenInformation]` <= (Datum + follow_up) ~ interval(Datum, `Date last seen[NTX PatientenInformation]`) / years(1),
    ## patient dropped after follow up
    `Date last seen[NTX PatientenInformation]` > (Datum + follow_up) ~ interval(Datum, (Datum + follow_up)) / years(1)
  )
)


data_ntx <- data_ntx %>%
  mutate(status = case_when(
    ## patient died within follow up
    `Todesdatum[NTX PatientenInformation]` <= (Datum + follow_up) ~ 1,
    ## patient died after follow up
    `Todesdatum[NTX PatientenInformation]` > (Datum + follow_up) ~ 0,
    ## patient dropped within follow up
    `Date last seen[NTX PatientenInformation]` <= (Datum + follow_up) ~ 0,
    ## patient dropped after follow up
    `Date last seen[NTX PatientenInformation]` > (Datum + follow_up) ~ 0
  )
)
```

```{r}
model_ntx_2 <- survfit(Surv(time = status_date, event = status,
                            type = "right") ~1, data = data_ntx)

ggsurvplot(model_ntx_2,
           conf.int = FALSE,
           cumevents = TRUE)
```

```{r}

```

```{r}

```

\newpage
# § 4 Cox regression

## § 4.1 IGA

- Event:
  - graft loss within follow up
- Censoring scheme:
  - if patient dropped within follow up, then censored by time dropped
  - if patient patient experienced graft loss after follow up, then censored by follow up end
  - if patient experienced death within follow up, then censored by death date else by follow up end
```{r}
data_iga <- data_iga %>% 
  ## censor/event date
  mutate(status_date = case_when(
    ## patient dropped during follow up
    (`T-dls` <= `T-date` + follow_up) ~ `T-dls`,
    ## patient experienced graft loss but after follow up
    ((!is.na(`graft loss date`)) &  (`graft loss date` > `T-date` + follow_up)) ~ `T-date` + follow_up,
    ## patient experienced graft loss within follow up
    ((!is.na(`graft loss date`)) &  (`graft loss date` <= `T-date` + follow_up)) ~ `graft loss date`,
    ## patient experienced no graft loss within follow up, neither dropped
    (is.na(`graft loss date`) & (`T-dls` > `T-date` + follow_up)) ~ `T-date` + follow_up
  )) %>% 
  ## status indicator
  mutate(status = case_when(
    ## patient dropped during follow up
    (`T-dls` <= `T-date` + follow_up) ~ 0,
    ## patient experienced graft loss but after follow up
    ((!is.na(`graft loss date`)) &  (`graft loss date` > `T-date` + follow_up)) ~ 0,
    ## patient experienced graft loss within follow up
    ((!is.na(`graft loss date`)) &  (`graft loss date` <= `T-date` + follow_up)) ~ 1,
    ## patient experienced no graft loss within follow up, neither dropped
    (is.na(`graft loss date`) & (`T-dls` > `T-date` + follow_up)) ~ 0
  ))
```

```{r}
data_iga <- data_iga %>%  
  ## censor/event date
  mutate(status_date = case_when(
    ## if graft loss within follow up
    `graft loss date` <= `T-date` + follow_up ~ `graft loss date`,
    ## else
    TRUE ~ `T-dls`
  )) %>% 
  ## censor/event indicator
  mutate(status = case_when(
    ## if graft loss within follow up
    data_iga$`graft loss date` <= data_iga$`T-date` + follow_up ~ 1,
    ## else
    TRUE ~ 0
  ))
```


```{r}
data_iga_pos <- data_iga[`biopsy proven recurrence (0=no, 1=yes)` == 1]
data_iga_neg <- data_iga[`biopsy proven recurrence (0=no, 1=yes)` == 0]
```

```{r}
model_iga_cox <- coxph(data = data_iga, formula = Surv(time = as.numeric(status_date),
                                      event = status) ~ R_age_Tdate +
                         data_iga$`R-sex` +
                         data_iga$`D-type` +
                         ## hla mismatch???
                         data_iga$`mm-A` +
                         data_iga$`Cold ischaemic period hours` +
                         data_iga$`D-age`)
summary(model_iga_cox)
```
```{r}
cox.zph(model_iga_cox)
## all p-values are relatively large, therefore the Null hypothesis of proportional hazards can not be rejected
```

## $ 4.2 NTX


```{r}
data_ntx <- data_ntx %>%
  mutate(status_date = case_when(
    ## patient died within follow up
    `Todesdatum[NTX PatientenInformation]` <= (Datum + follow_up) ~ interval(Datum, `Todesdatum[NTX PatientenInformation]`) / years(1),
    ## patient died after follow up
    `Todesdatum[NTX PatientenInformation]` > (Datum + follow_up) ~ interval(Datum, Datum + follow_up) / years(1),
    ## patient dropped within follow up
    `Date last seen[NTX PatientenInformation]` <= (Datum + follow_up) ~ interval(Datum, `Date last seen[NTX PatientenInformation]`) / years(1),
    ## patient dropped after follow up
    `Date last seen[NTX PatientenInformation]` > (Datum + follow_up) ~ interval(Datum, Datum + follow_up) / years(1)
  )
)

data_ntx <- data_ntx %>%
  mutate(status = case_when(
    ## patient died within follow up
    `Todesdatum[NTX PatientenInformation]` <= (Datum + follow_up) ~ 1,
    ## patient died after follow up
    `Todesdatum[NTX PatientenInformation]` > (Datum + follow_up) ~ 0,
    ## patient dropped within follow up
    `Date last seen[NTX PatientenInformation]` <= (Datum + follow_up) ~ 0,
    ## patient dropped after follow up
    `Date last seen[NTX PatientenInformation]` > (Datum + follow_up) ~ 0
    
  )
)
```

```{r}
model_ntx_cox <- coxph(formula = Surv(time = status_date, event = status) ~ R_age_Datum +
                         Geschlecht + `TX Status[NTX PatientenInformation]`,
                       data = data_ntx)

summary(model_ntx_cox)
```

```{r}
cox.zph(model_ntx_cox)
## all p-values are relatively large, therefore the Null hypothesis of proportional hazards can not be rejected
```